---
title: "Analysis of Activity Monitoring Data"
author: "Pallak Goyal"
date: "2023-01-19"
output: html_document
---
## Loading and processing the data
The data set was first loaded into R.
```{r, loaddata}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url,"./activity.data.zip")
unzip("./activity.data.zip")
activity.data <- read.csv("./activity.csv")
str(activity.data)
```
The data was cleaned to make analysis possible.
```{r, cleandata}
activity.data$date <- as.Date(activity.data$date, format = "%Y-%m-%d")
str(activity.data)
```
## What is the mean of the total number of steps taken per day?
The data set of total number of steps per day was created.
```{r data}
steps.per.day <- aggregate.data.frame(activity.data$steps, by = list(date = activity.data$date), FUN = sum)
names(steps.per.day)= c("date", "total.steps")
str(steps.per.day)
```
The number of steps taken each day is represented as a histogram.
```{r, hist}
library(ggplot2)
qplot(x = steps.per.day$total.steps, xlab = "Total Steps taken in a Day", ylab = "Frequency", main = "Histogram of Total Steps taken in a day")
```

The mean number of steps taken per day was calculated
```{r, mean}
mean(steps.per.day$total.steps, na.rm = TRUE)
```
The median number of steps taken per day was calculated
```{r, median}
median(steps.per.day$total.steps, na.rm = TRUE)
```
## What is the average daily activity pattern?
The data set of the average steps across days per time interval was created.
```{r, data2}
daily.activity <- aggregate(activity.data$steps, by = list(interval = activity.data$interval), FUN = mean, na.rm = TRUE)
names(daily.activity) <- c("interval", "mean.steps")
str(daily.activity)
```
A time series plot at 5 minutes interval was created.
```{r, timeplot}
plot(x = daily.activity$interval, y = daily.activity$mean.steps, type = "l", main = "Time Plot of Average daily activity", xlab = "Time Interval", ylab = "Mean Steps")
```
The greatest amount of activity is observed between 750 and 1000 minutes into the day.There is very little activity between 0 to 500 minutes into the day.
The following code was run to find the interval with maximum average footsteps.
```{r, maxavg}
daily.activity$interval[which.max(daily.activity$mean.steps)]
```
## Imputing Missing Values
The following code returns the total number of rows in the main data set that contain missing values.
```{r, missing}
nrow(activity.data[activity.data$steps == "NA", ])
```
The missing data should be replaced with the median value for all the days in that time interval.The following code creates a new data set with missing values filled in. As suggested earlier, the interval median values are used to fill in the missing data.
```{r, replace.na}
activity.data2 <- activity.data
for (i in 1: nrow(activity.data2)){
        if(is.na(activity.data2[i,1])){
                activity.data2[i,1] <- daily.activity[match(activity.data2[i,3], daily.activity$interval),2]
        }
} 
str(activity.data2)
```
The data set of the total number of steps per day was created using the data without missing values.
```{r}
steps.per.day2 <- aggregate.data.frame(activity.data2$steps, by = list(date = activity.data2$date), FUN = sum)
names(steps.per.day2)= c("date", "total.steps")
str(steps.per.day2)
```
The number of steps taken each day is represented as a histogram.
```{r}
library(ggplot2)
qplot(x = steps.per.day2$total.steps, xlab = "Total Steps taken in a Day", ylab = "Frequency", main = "Histogram of Total Steps taken in a day")
```
The mean number of steps taken per day was calculated
```{r}
mean(steps.per.day2$total.steps, na.rm = TRUE)
```
The median number of steps taken per day was calculated
```{r}
median(steps.per.day2$total.steps, na.rm = TRUE)
```
As can be seen by comparing with the previous results there is not much change after replacing the missing values with the median.Although the imputing of missing values has increased the total daily number of steps, it does not affect the results because the missing values had been removed to calculate the earlier results.
## Are there differences in activity patterns between weekdays and weekends?
The following code creates the data frame with an additional column that identifies whether the day is a weekday or weekend.
```{r weekday}
activity.data2$date <- as.Date(activity.data2$date, format = "%Y-%m-%d")
activity.data2$day <- weekdays(activity.data2$date)
activity.data2$day.type <- c(rep(0, nrow(activity.data2)))
 for (i in 1: nrow(activity.data2)){
        if (activity.data2[i, ]$day == "Sunday"| activity.data2[i, ]$day == "Saturday"){
        activity.data2[i,]$day.type <- "weekend"
        }
        else{
                activity.data2[i, ]$day.type <- "weekday"
        }
}
activity.data2$day.type <- as.factor(activity.data2$day.type)
str(activity.data2)
```
The following code creates the data frame with average daily activty across intervals for weekdays and weekends.
```{r}
daily.activity2 <- aggregate(activity.data2$steps,
                             by = list(activity.data2$interval, 
                                       activity.data2$day.type), FUN = mean)
names(daily.activity2) <- c("interval", "day.type","mean.steps")
str(daily.activity2)
```
The following code creates a panel time plot for average daily activiy.
```{r}
library(lattice)
xyplot(mean.steps~interval|day.type, data = daily.activity2, type = "l", xlab = "Interval", ylab="Average Steps", main= "Time Plot by Day Type")
```
